Resource Adequacy and Operational Security Interaction in the EPOC 2030-50 Project

### 2022-11-22 papers/main.tex
Saving

@@

Secondly, the adequacy assessment is done using an unconventional \ac{UC} model, a modified version of the \ac{DUCPR} model described in \cite{Bruninx2017}. This model can be seen as a compromise between a computationally expensive stochastic \ac{UC} model and a more tractable though less accurate deterministic \ac{UC} model with operating reserve requirements. The treatment of operating reserves in this model allows a trade-off between shedding load in the day ahead stage, which would reduce day ahead adequacy, and shedding operating reserves, which should reduce operational security. By making this trade-off explicit and verifying the results in an operational security analysis we aim to elaborate on the results of \cite{Hermans2018}, which did not consider the possibility that shedding operating reserves would lead to insecure operation of the electric power system.

@@

\begin{figure}[ht]
    \centering
    % \includegraphics[width=0.9\textwidth]{../plots/nodal_imbalance_investigation/imbalance_range}
    \caption{Range of possible imbalances for all hours of day 309.\label{fig:imbalance_range}}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    % \includegraphics[width=0.9\textwidth]{../plots/nodal_imbalance_investigation/convex_hull_TIHANGE 1_TIHANGE 2}
    \caption{Illustration of convex hull constraints on $d_{nlt}^{L^+}$ and $d_{nlt}^{L^-}$. Blue circles are imbalance scenarios, shaded area is the convex hull, $d_{nt}^{min}$ and $d_{nt}^{max}$ are the x limits for TIHANGE 1 and the y limits for TIHANGE 2. \label{fig:convex_hull_tihange}}
\end{figure}

@@

The grid and resource data used for this interaction is inspired by the case study in \cite{Belderbos2020}, omitting the gas network and power to gas technologies. This data gives a stylised Belgian system with a high penetration (~8s0\%) of \ac{VRES}, which includes 75.7 GW of solar PV, 7.0 GW of Onshore Wind and 7.3 GW of Offshore Wind. The grid consists of 46 nodes connected by 69 lines. The total amount of conventional thermal generation is 9.9 GW, with all generators based on new \ac{CCGT} units apart from one new \ac{OCGT} unit at Lixhe. The nuclear generators of Doel and Tihange are therefore replaced with new \ac{CCGT} units.

Power to Gas is omitted but 14.1 GW of storage power capacity (1.1 GW provided by pumped hydro at Coo, the rest batteries) and 101 GWh of storage energy capacity (8.2 GWh provided by pumped hydro at Coo, the rest batteries) is included. The batteries therefore have a duration of 7.2 hours (which could be considered atypical, given current durations of 1 - 2 hours \addcite{MIT paper for this?}).

The network topology was obtained from the Belgian TSO, Elia \cite{Elia2018grid}. It consists of 46 nodes connected by 69 lines.

The load time series used was the load as seen by Elia on the high-voltage grid from the year 2015 \todo{add cite.}. For this year, the minimum and maximum load was 5,777 MW and 13,670 MW respectively, with an average of 9,934 MW. The electrical load is assigned to the different nodes in the system according to the rated capacity of the transformer feeding each node.

The \ac{VRES} availability factor time series are taken from Elia data for the year 2015, same as the load. This was done by dividing generation from offshore wind, wind and solar by the total installed capacities available. This allows scaling up the generation of \ac{VRES} to the previously mentioned capacities. The distribution of onshore wind and solar photovoltaic (PV) capacity was done according to the distribution made in \cite{DelarueErik2016Dtio}.

@@

A complete overview of the data is given in \cite{Belderbos2020}, Appendix B.

\subsubsection{Methodology used to select days for analysis}

@@

\begin{table}[H]
    \centering
    \begin{tabular}{cccc}
        \toprule
        Day ahead daily load shedding [MWh/day] & Day of year & Month & Day of month \\
        \midrule
        0                                       & 161         & 6     & 10           \\
        449                                     & 319         & 11    & 15           \\
        1,247                                   & 285         & 10    & 12           \\
        10,062                                  & 12          & 1     & 12           \\
        \bottomrule
    \end{tabular}
    \caption{Days selected for analysis. These were selected based on the results of a full year economic dispatch model with load multiplied uniformly across all nodes by a factor of 1.5. The 0, 33, 66 and 100th percentiles in terms of daily load shedding were then selected.}\label{tab:day_selection}
\end{table}

@@

\begin{table}[H]
    \centering
    \begin{tabular}{ccccc}
        \toprule
        UC  & DANet & PSCD & Load shedding [MWh] & Objective [\euro] \\
        \midrule
            &       &      & 0.0                 & 3,127,814         \\
        \xm &       &      & 0.0                 & 4,003,654         \\
        \xm & \xm   &      & 0.0                 & 4,394,693         \\
        \xm & \xm   & \xm  & 0.0                 & 4,394,513         \\
        \bottomrule
    \end{tabular}
    \caption{Analysis of model results when operating reserves are not included. UC = Unit Commitment constraints, DANet = Day Ahead Network constraints, PSCD = Prevent Simultaneous Charging and Discharging constraints.}\label{tab:results_no_OR}
\end{table}

@@

\todo[inline]{Confirm the above suspicion}

\begin{figure}[H]I
    \centering
    \includegraphics[width=0.7\textwidth]{../data/sims/main_model_runs/285/base_UC=true_DANet=true_RSV=1.0/dispatch}
    \caption{Dispatch schedule for day 285, unit commitment constraints, day ahead network constraints and operating reserve requirements but no limit on reserve shedding.\label{fig:dispatch_with_operating_reserves}}
\end{figure}

Considering the case with just unit commitment constraints, the reserve shedding limit can be decreased without increasing load shedding, though there is an increase (~0.09\%) in the objective function greater than the MIP gap. This is counterintuitive, as you would expect the DUC-PR model to shed all operating reserves before shedding load in the day ahead stage. The former is less costly than the latter after all, due to activation probabilities being < 1. However, what is occuring is most likely `economic reserve shedding' - shedding reserves because they are highly unlikely to be activated. This phenomenon was also observed in \cite{Hermans2018}. \figref{fig:reserves_with_operating_reserves} illustrates that almost all reserve shedding occurs between 14:00 and 16:00 (hour 6830 to 6832) and not 8:00 to 10:00 as with load shedding. These are hours with a lot of solar generation and very little conventional, so to avoid this reserve shedding only requires additional generator commitment and not increasing the amount of load shedding in those hours.

@@

\begin{table}[ht]
    \centering
    \footnotesize
    \begin{tabular}{ccccccccc}
        \toprule
        UC  & DANet & PSCD & RANet & AbsImb & RSL & Load shedding [MWh] & Reserve Shedding [MWh] & Objective [\euro] (MIP Gap [\%]) \\
        \midrule
        \xm & \xm   & \xm  &       &        & 0.0 & 5,410               & 0.0                    & 63,198,858                       \\
        \xm & \xm   & \xm  &       &        & 0.5 & 5,410               & 5,112                  & 63,139,965                       \\
        \xm & \xm   & \xm  &       &        & 1.0 & 5,410               & 5,114                  & 63,140,337                       \\
        \midrule
        \xm & \xm   & \xm  & \xm   &        & 0.0 & 5,410               & 0                      & 63,199,116                       \\
        \xm & \xm   & \xm  & \xm   &        & 0.5 & 5,410               & 4,898                  & 63,139,874                       \\
        \xm & \xm   & \xm  & \xm   &        & 1.0 & 5,410               & 5,611                  & 63,137,767                       \\
        \midrule
        \xm & \xm   & \xm  & \xm   & \xm    & 0.0 & 5,410               & 0                      & 63,232,938 (0.04)
        )                                                                                                                           \\
        \xm & \xm   & \xm  & \xm   & \xm    & 0.5 & 5,410               & 5,687                  & 63,150,183                       \\
        \xm & \xm   & \xm  & \xm   & \xm    & 1.0 & 5,410               & 6,452                  & 63,155,353 (0.02)                \\
        \bottomrule
    \end{tabular}
    \caption{Analysis of model results with operating reserves, with and without reserve activation network constraints and with and without absolute limits on nodal imbalances. MIP gap is shown in the case where this was greater than 0.01\%. UC = Unit Commitment constraints, DANet = Day Ahead Network constraints, PSCD = Prevent Simultaneous Charging and Discharging constraints, RANet = Reserve Activation Network constraints, AbsImb = Absolute limits on nodal Imbalance realisation, RSL = Reserve Shedding Limit.}\label{tab:results_with_RANet}
\end{table}

@@

\begin{table}[ht]
    \centering
    \footnotesize
    \begin{tabular}{ccccccccc}
        \toprule
        UC  & DANet & PSCD & RANet & AbsImb & RSL & Load shedding [MWh] & Reserve Shedding [MWh] & Objective [\euro] (MIP Gap [\%]) \\
        \midrule
            &       &      & \xm   &        & 0.0 & 5,825               & 0.0                    & 69,902,500                       \\
        \xm &       &      & \xm   &        & 0.0 & 5,825               & 0.0                    & 77,491,233                       \\
        \xm & \xm   &      & \xm   &        & 0.0 & 6,314               & 0.0                    & 82,523,020                       \\
        \xm & \xm   & \xm  & \xm   &        & 0.0 & 6,314               & 0.0                    & 82,523,020                       \\
        \midrule
        \xm & \xm   & \xm  & \xm   &        & 0.0 & 6,314               & 0.0                    & 82,523,020                       \\
        \xm & \xm   & \xm  & \xm   &        & 0.5 & 5,410               & 12,187                 & 72,771,511                       \\
        \xm & \xm   & \xm  & \xm   &        & 1.0 & 5,410               & 12,244                 & 72,771,897                       \\
        \midrule
        \xm & \xm   & \xm  & \xm   & \xm    & 0.0 & 6,314               & 0                      & 83,163,868                       \\
        \xm & \xm   & \xm  & \xm   & \xm    & 0.5 & 5,410               & 13,115                 & 73,298,008                       \\
        \xm & \xm   & \xm  & \xm   & \xm    & 1.0 & 5,410               & 13,430                 & 73,296,805                       \\
        \bottomrule
    \end{tabular}
    \caption{Analysis of model results with operating reserves, with and without reserve activation network constraints and with and without absolute limits on nodal imbalances and for load multiplied by a factor of 1.5. MIP gap is shown in the case where this was greater than 0.01\%. UC = Unit Commitment constraints, DANet = Day Ahead Network constraints, PSCD = Prevent Simultaneous Charging and Discharging constraints, RANet = Reserve Activation Network constraints, AbsImb = Absolute limits on nodal Imbalance realisation, RSL = Reserve Shedding Limit.}\label{tab:results_with_RANet_LM_15}
\end{table}

@@

>>>>

\begin{abstract}
    To be filled in
\end{abstract}

@@

Secondly, the adequacy assessment is done using an unconventional \ac{UC} model, a modified version of the \ac{DUCPR} model described in~\cite{Bruninx2017}. This model can be seen as a compromise between a computationally expensive stochastic \ac{UC} model and a more tractable though less accurate deterministic \ac{UC} model with operating reserve requirements. The treatment of operating reserves in this model allows a trade-off between shedding load in the day ahead stage, which would reduce day ahead adequacy, and shedding operating reserves, which should reduce operational security. By making this trade-off explicit and verifying the results in an operational security analysis we aim to elaborate on the results of~\cite{Hermans2018}, which did not consider the possibility that shedding operating reserves would lead to insecure operation of the electric power system.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.9\textwidth]{../plots/nodal_imbalance_investigation/imbalance_range}
    \caption{Range of possible imbalances for all hours of day 309.\label{fig:imbalance_range}}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.9\textwidth]{../plots/nodal_imbalance_investigation/convex_hull_TIHANGE 1_TIHANGE 2}
    \caption{Illustration of convex hull constraints on $d_{nlt}^{L^+}$ and $d_{nlt}^{L^-}$. Blue circles are imbalance scenarios, shaded area is the convex hull, $d_{nt}^{min}$ and $d_{nt}^{max}$ are the x limits for TIHANGE 1 and the y limits for TIHANGE 2. \label{fig:convex_hull_tihange}}
\end{figure}

@@

The grid and resource data used for this interaction is inspired by the case study in~\cite{Belderbos2020}, omitting the gas network and power to gas technologies. This data gives a stylised Belgian system with a high penetration (~8s0\%) of \ac{VRES}, which includes 75.7 GW of solar PV, 7.0 GW of Onshore Wind and 7.3 GW of Offshore Wind. The grid consists of 46 nodes connected by 69 lines. The total amount of conventional thermal generation is 9.9 GW, with all generators based on new \ac{CCGT} units apart from one new \ac{OCGT} unit at Lixhe. The nuclear generators of Doel and Tihange are therefore replaced with new \ac{CCGT} units.

Power to Gas is omitted but 14.1 GW of storage power capacity (1.1 GW provided by pumped hydro at Coo, the rest batteries) and 101 GWh of storage energy capacity (8.2 GWh provided by pumped hydro at Coo, the rest batteries) is included. The batteries therefore have a duration of 7.2 hours (which could be considered atypical, given current durations of 1 - 2 hours~\cite{MITEI2022}).

The network topology was obtained from the Belgian TSO, Elia~\cite{Elia2018grid}. It consists of 46 nodes connected by 69 lines.

The load time series used was the load as seen by Elia on the high-voltage grid from the year 2015~\cite{Belderbos2020}. For this year, the minimum and maximum load was 5,777 MW and 13,670 MW respectively, with an average of 9,934 MW. The electrical load is assigned to the different nodes in the system according to the rated capacity of the transformer feeding each node.

The \ac{VRES} availability factor time series are taken from Elia data for the year 2015, same as the load. This was done by dividing generation from offshore wind, wind and solar by the total installed capacities available. This allows scaling up the generation of \ac{VRES} to the previously mentioned capacities. The distribution of onshore wind and solar photovoltaic (PV) capacity was done according to the distribution made in~\cite{DelarueErik2016Dtio}.

@@

A complete overview of the data is given in~\cite{Belderbos2020}, Appendix B.

@@

\begin{table}[ht]
    \centering
    \begin{tabular}{cccc}
        \toprule
        Day ahead daily load shedding [MWh/day] & Day of year & Month & Day of month \\
        \midrule
        0                                       & 161         & 6     & 10           \\
        449                                     & 319         & 11    & 15           \\
        1,247                                   & 285         & 10    & 12           \\
        10,062                                  & 12          & 1     & 12           \\
        \bottomrule
    \end{tabular}
    \caption{Days selected for analysis. These were selected based on the results of a full year economic dispatch model with load multiplied uniformly across all nodes by a factor of 1.5. The 0, 33, 66 and 100th percentiles in terms of daily load shedding were then selected.}\label{tab:day_selection}
\end{table}

@@

\begin{table}[ht]
    \centering
    \begin{tabular}{ccccc}
        \toprule
        UC  & DANet & PSCD & Load shedding [MWh] & Objective [\euro] \\
        \midrule
            &       &      & 0.0                 & 3,127,814         \\
        \xm &       &      & 0.0                 & 4,003,654         \\
        \xm & \xm   &      & 0.0                 & 4,394,693         \\
        \xm & \xm   & \xm  & 0.0                 & 4,394,513         \\
        \bottomrule
    \end{tabular}
    \caption{Analysis of model results when operating reserves are not included. UC = Unit Commitment constraints, DANet = Day Ahead Network constraints, PSCD = Prevent Simultaneous Charging and Discharging constraints.}\label{tab:results_no_OR}
\end{table}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\textwidth]{../data/sims/main_model_runs/285/base_UC=true_DANet=true_RSV=1.0/dispatch}
    \caption{Dispatch schedule for day 285, unit commitment constraints, day ahead network constraints and operating reserve requirements but no limit on reserve shedding.\label{fig:dispatch_with_operating_reserves}}
\end{figure}

Considering the case with just unit commitment constraints, the reserve shedding limit can be decreased without increasing load shedding, though there is an increase (~0.09\%) in the objective function greater than the MIP gap. This is counterintuitive, as you would expect the DUC-PR model to shed all operating reserves before shedding load in the day ahead stage. The former is less costly than the latter after all, due to activation probabilities being < 1. However, what is occuring is most likely `economic reserve shedding' - shedding reserves because they are highly unlikely to be activated. This phenomenon was also observed in~\cite{Hermans2018}. \figref{fig:reserves_with_operating_reserves} illustrates that almost all reserve shedding occurs between 14:00 and 16:00 (hour 6830 to 6832) and not 8:00 to 10:00 as with load shedding. These are hours with a lot of solar generation and very little conventional, so to avoid this reserve shedding only requires additional generator commitment and not increasing the amount of load shedding in those hours.

@@

\begin{table}[ht]
    \centering
    \tiny
    \begin{tabular}{ccccccccc}
        \toprule
        UC  & DANet & PSCD & RANet & AbsImb & RSL & Load shedding [MWh] & Reserve Shedding [MWh] & Objective [\euro] (MIP Gap [\%]) \\
        \midrule
        \xm & \xm   & \xm  &       &        & 0.0 & 5,410               & 0.0                    & 63,198,858                       \\
        \xm & \xm   & \xm  &       &        & 0.5 & 5,410               & 5,112                  & 63,139,965                       \\
        \xm & \xm   & \xm  &       &        & 1.0 & 5,410               & 5,114                  & 63,140,337                       \\
        \midrule
        \xm & \xm   & \xm  & \xm   &        & 0.0 & 5,410               & 0                      & 63,199,116                       \\
        \xm & \xm   & \xm  & \xm   &        & 0.5 & 5,410               & 4,898                  & 63,139,874                       \\
        \xm & \xm   & \xm  & \xm   &        & 1.0 & 5,410               & 5,611                  & 63,137,767                       \\
        \midrule
        \xm & \xm   & \xm  & \xm   & \xm    & 0.0 & 5,410               & 0                      & 63,232,938 (0.04)
        )                                                                                                                           \\
        \xm & \xm   & \xm  & \xm   & \xm    & 0.5 & 5,410               & 5,687                  & 63,150,183                       \\
        \xm & \xm   & \xm  & \xm   & \xm    & 1.0 & 5,410               & 6,452                  & 63,155,353 (0.02)                \\
        \bottomrule
    \end{tabular}
    \caption{Analysis of model results with operating reserves, with and without reserve activation network constraints and with and without absolute limits on nodal imbalances. MIP gap is shown in the case where this was greater than 0.01\%. UC = Unit Commitment constraints, DANet = Day Ahead Network constraints, PSCD = Prevent Simultaneous Charging and Discharging constraints, RANet = Reserve Activation Network constraints, AbsImb = Absolute limits on nodal Imbalance realisation, RSL = Reserve Shedding Limit.}\label{tab:results_with_RANet}
\end{table}

@@

\begin{table}[ht]
    \centering
    \tiny
    \begin{tabular}{ccccccccc}
        \toprule
        UC  & DANet & PSCD & RANet & AbsImb & RSL & Load shedding [MWh] & Reserve Shedding [MWh] & Objective [\euro] (MIP Gap [\%]) \\
        \midrule
            &       &      & \xm   &        & 0.0 & 5,825               & 0.0                    & 69,902,500                       \\
        \xm &       &      & \xm   &        & 0.0 & 5,825               & 0.0                    & 77,491,233                       \\
        \xm & \xm   &      & \xm   &        & 0.0 & 6,314               & 0.0                    & 82,523,020                       \\
        \xm & \xm   & \xm  & \xm   &        & 0.0 & 6,314               & 0.0                    & 82,523,020                       \\
        \midrule
        \xm & \xm   & \xm  & \xm   &        & 0.0 & 6,314               & 0.0                    & 82,523,020                       \\
        \xm & \xm   & \xm  & \xm   &        & 0.5 & 5,410               & 12,187                 & 72,771,511                       \\
        \xm & \xm   & \xm  & \xm   &        & 1.0 & 5,410               & 12,244                 & 72,771,897                       \\
        \midrule
        \xm & \xm   & \xm  & \xm   & \xm    & 0.0 & 6,314               & 0                      & 83,163,868                       \\
        \xm & \xm   & \xm  & \xm   & \xm    & 0.5 & 5,410               & 13,115                 & 73,298,008                       \\
        \xm & \xm   & \xm  & \xm   & \xm    & 1.0 & 5,410               & 13,430                 & 73,296,805                       \\
        \bottomrule
    \end{tabular}
    \caption{Analysis of model results with operating reserves, with and without reserve activation network constraints and with and without absolute limits on nodal imbalances and for load multiplied by a factor of 1.5. MIP gap is shown in the case where this was greater than 0.01\%. UC = Unit Commitment constraints, DANet = Day Ahead Network constraints, PSCD = Prevent Simultaneous Charging and Discharging constraints, RANet = Reserve Activation Network constraints, AbsImb = Absolute limits on nodal Imbalance realisation, RSL = Reserve Shedding Limit.}\label{tab:results_with_RANet_LM_15}
\end{table}

@@

\subsection{Simple illustratve case}

We begin by illustrating the tradeoff that exists between real time load shedding and day ahead reserve shedding with a simple model. No network or unit commitment constraints are included and no storage either. The latter choice is made since in this model formulation storage is not able to contribute to operating reserve requirements, complicating the interpretation of the results. Day 319 is chosen since resource use is not as `tight' as the other days and the absence of storage increases the probability of load or reserve shedding.

A comparison of the real time operational security from the DUC-PR model compared to those obtained from a power flow performed by the team ULiege is given in the figure below.
